A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-31 latex/main.tex
Updates from Overleaf

%%% WIP: SHZ
We present the first pandemic-scale
ancestral recombination graph (ARG) for SARS-CoV-2,
reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
We demonstrate how this ARG, for the first time,
enables systematic identification and characterization of
recombination events in fine-grained
detail--in terms of their parents, breakpoints, and timing of
emergence--providing richer and more nuanced
genealogical, genomic, and epidemiological context
than existing phylogenetic approaches.
Together, sc2ts and the ARG represent a new paradigm in
how we study the evolution of pathogens in the face of ongoing sampling.

Accounting for recombination is a key challenge in pathogen genomics, and
recombination has been largely ignored in analysis of viral genomes.
Existing phylogenetic approaches for pathogens detect recombination
\textit{post hoc} (e.g., UShER+RIPPLES),
treat it as a nuance (e.g., XX), or
discard its signals from downstream analyses (e.g., XX).
Sc2ts addresses this challenge
by using the ARG as the underlying data model,
explicitly representing recombination events as individual nodes and
relating them to specific parents in the ARG.
The ARG provides a coherent framework to determine and represent
the relationships between recombinants and their specific parents
(which are not possible using a tree-based data model,
e.g., the Mutation Annotated Tree format of UShER).
The ARG reveals and clarifies the relationships between
individual samples of the same recombinant origin,
avoiding issues such as over-naming recombinants
(e.g., the samples assigned to
five differently labeled recombinant Pango lineages
descending from the “Xx” recombination event in the ARG).
Our results show how an ARG can considerably simplify
efforts to identify and monitor emerging recombinants,
facilitating communication of genomic surveillance results.

% SHZ: A bit vague, but is this worth discussing?
% What else to say about the timing of these events?
% TODO: Example with say XBB?
% TODO: Say we don’t explore this fully, but it could be a follow-up study?
The timing of recombination events is of key interest in
pathogen genomic epidemiology,
because this information may yield insights into
% Probably A bad idea to say this.
how the emergence of a new recombinant might be
linked with changes in public health policies.
Using the ARG, we reconstructed a historical timeline of
recombination events during the COVID-19 pandemic,
showing that it is possible to pinpoint the occurrence of
recombination events precisely in time,
within windows of X to X months
(given the intensity of sampling in the COVID-19 dataset).

@@

The ARG itself is a powerful resource
to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies
important features of SARS-CoV-2 evolution,
which are shown in different studies
using various methods, tools, and approaches.
The ARG also compactly encodes genetic variation in
millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of
this valuable dataset for a range of
genetic and evolutionary analyses beyond those presented here,
for instance, to understand
how the variability in sequencing efforts
across time and location affects our ability
to reconstruct the origin of important recombinant lineages and
to estimate when they emerge.
Because building the ARG is computationally expensive, and
checking and improving various aspects of its quality is laborious,
we hope that making this resource openly available will
save significant time and effort for other researchers,
enabling immediate focus on analyzing the information
contained within the COVID-19 dataset.
% TODO: Briefly discussion software tools here or in a subsequent paragraph.

>>>>

%%% SHZ: WIP alternative.
We present the first pandemic-scale
ancestral recombination graph (ARG) for SARS-CoV-2,
reconstructed from $\sim$2.48 million genomes using \textit{sc2ts}.
We demonstrate how this ARG, for the first time,
enables systematic identification and characterization of
recombination events in fine-grained
detail--in terms of their parents, breakpoints, and timing of
emergence--providing richer and more nuanced
genealogical, genomic, and epidemiological context
than existing phylogenetic approaches.
Together, sc2ts and the ARG represent a new paradigm in
how we study the evolution of pathogens in the face of ongoing sampling.

Accounting for recombination is a key challenge in pathogen genomics, and
recombination has been largely ignored in analysis of viral genomes.
Existing phylogenetic approaches for pathogens detect recombination
\textit{post hoc} (e.g., UShER+RIPPLES),
treat it as a nuance (e.g., XX), or
discard its signals from downstream analyses (e.g., XX).
Sc2ts addresses this challenge
by using the ARG as the underlying data model,
explicitly representing recombination events as individual nodes and
relating them to specific parents in the ARG.
The ARG provides a coherent framework to determine and represent
the relationships between recombinants and their specific parents
(which are not possible using a tree-based data model, e.g., the MAT format of UShER).
The ARG reveals and clarifies the relationships between
individual samples of the same recombinant origin,
avoiding issues such as over-naming recombinants
(e.g., the samples assigned to
five differently labeled recombinant Pango lineages
descending from the “Xx” recombination event in the ARG).
Our results show how an ARG can considerably simplify
efforts to identify and monitor emerging recombinants,
facilitating communication of genomic surveillance results.

% SHZ: Very WIP
% Discuss the new preprint from Turakhia.
The timing of recombination events is of key interest in
pathogen genomic epidemiology.
because this information may yield insights into
how the emergence of a new recombinant might be
linked with changes in public health policies.

@@

The ARG itself is a powerful resource
to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies
important features of SARS-CoV-2 evolution,
which are shown in different studies
using various methods, tools, and approaches.
The ARG also compactly encodes genetic variation in
millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of
this valuable dataset for a range of
genetic and evolutionary analyses beyond those presented here,
for instance, to understand
how the variability in sequencing efforts
across time and location affects our ability
to reconstruct the origin of important recombinant lineages and
to estimate when they emerge.
Because building the ARG is computationally expensive, and
checking and improving various aspects of its quality is laborious,
we hope that making this resource openly available will
save significant time and effort for other researchers,
enabling immediate focus on analyzing the information
contained within the COVID-19 dataset.

% TODO: Briefly discussion software tools here or in a subsequent paragraph.
